---
title: "Air Pollution"
author: "bdorhan"
output:
  html_document:
    df_print: paged
---

The next chunk contains the function that calculates the mean of a pollutant (sulfate or nitrate) across a specified list of monitors.The function 'pollutantmean' takes three arguments: 'directory', 'pollutant', and 'id'. Given a vector monitor ID numbers, 'pollutantmean' reads that monitors' particulate matter data from the directory specified in the 'directory' argument and returns the mean of the pollutant across all of the monitors, ignoring any missing values coded as NA.

```{r pollutantmean, echo=TRUE, warning=TRUE}
pollutantmean <- function(directory, pollutant, id = 1:332) {
  ## 'directory' is a character vector of length 1 indicating the location of the CSV files
  ## 'polutant' is the chacter vector of length 1 indicating the name of the pollutant for which we will calculate the        mean; either "sulfate" or "nitrate"
  ## 'id' is an integer vector indicating the monitor ID numbers to be used
  ## Returns the mean of the pollutant across the monitor ID numbers
  
  files <- list.files(path = directory, pattern = ".csv", full.names = TRUE)
  values <- numeric()
  
  for (i in id) {
    data <- read.csv(files[i])
    values <- c(values, data[[pollutant]])
  }
  
  mean(values, na.rm = TRUE)
 }
```
```{r pollutmean_run, echo=TRUE, warning=TRUE}
pollutantmean("specdata", "nitrate")
```

The next chunk contains the function that reads a directory full of files and reports the number of completely observed cases in each data file.

```{r complete, echo=TRUE, warning=TRUE}
complete <- function(directory, id = 1:332) {
  ## 'directory' is a character vector of length 1 indicating the location of the CSV files
  ## 'id' is an integer vector indicating the monitor ID numbers to be used
  ## Returns a dataframe of the form:
  ## id nobs
  ## 1  117
  ## 2  1041
  ## ...
  ## where 'id' is the monitor ID number and 'nobs' is the number of complete cases
  
  files <- list.files(path = directory, pattern = ".csv", full.names = TRUE)
  print(paste("##", "", "id", "nobs"))
  counter <- 0
  for (i in id) {
    counter <- counter + 1
    data = read.csv(files[i])
    print(paste("##", counter, i, sum(complete.cases(data))))
  }
}
```


```{r complete_run, echo=TRUE, warning=TRUE}
complete("specdata", 54)
```
The last chunk contains the function that takes a directory of data files and a threshold for complete cases and calculates the correlation between sulfate and nitrate for monitor locations where the number of completely observed cases (on all variables) is greater than the threshold.
```{r corr, echo=TRUE, warning=TRUE}
corr <- function(directory, threshold = 0) {
  ## 'directory' is a character vector of length 1 indicating the location of the CSV files
  ## 'threshold' is a numeric vector of length 1 indicating the number of comletely observed observations required to      compute the correlation between nitrate and sulfate; default value is 0
  ## Returns a numeric vector of correlations
  
  files <- list.files(path = directory, pattern = ".csv", full.names = TRUE)
  cors <- numeric()
  for (i in 1:332) {
    data <- read.csv(files[i])
    if (sum(complete.cases(data)) > threshold) {
      cors <- c(cors, cor(data[["sulfate"]], data[["nitrate"]], use = "complete.obs"))
    }
  }
  cors
}
```


```{r corr_run, echo=TRUE, warning=TRUE}
cr <- corr("specdata", 150)

head(cr)

summary(cr)

length(cr)
```

